---
title: "BoskR Tutorial"
author: "[Orlando Schwery](https://oschwery.github.io/)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BoskR Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1 Introduction ##
This is a tutorial vignette to provide a quick and easy introduction to the use of BoskR. It will make use of the inbuilt example files and lead through the most important functions of the package.

The purpose of BoskR is, to test whether a given diversification model is adequate to analyse an empirical phylogeny with. We can use different models to analyse diversification on a phylogeny, and pick the one that's best supported ( _e.g._ by likelihood or AIC), but that doesn't tell us anything about whether that model is an appropriate one to use in the first place, _i.e._ whether the model is capable of describing the patterns we are attempting to analyse at all.

The way BoskR is trying to judge adequacy is by tree shape. The logic behind this is, that if a model cannot produce trees of a shape similar to an empirical tree, it is also not suited to analyse that tree. The package makes use of tree metrics measured from the spectral density of a tree's modified graph Laplacian, as obtained through the package RPANDA. Details on this can be found in the corresponding publication^[Eric Lewitus, Helene Morlon; Characterizing and Comparing Phylogenies from their Laplacian Spectrum, Systematic Biology, Volume 65, Issue 3, 1 May 2016, Pages 495–507, [https://doi.org/10.1093/sysbio/syv116](https://doi.org/10.1093/sysbio/syv116)] and package documentations^[[Package RPANDA on CRAN](https://CRAN.R-project.org/package=RPANDA)].

These metrics have now also been complemented with more 'classical' metrics (_e.g._, Colless, node ages), to gain more information on the trees (and perhaps save some time) - see **section 3.6**.

## 2 Installation ##
### 2.1 CRAN ###
To just use the stable out of the box version of BoskR, install the stable CRAN version of the package:

```{r, eval=FALSE}
install.packages("BoskR")
```

### 2.2 Development Version from GitHub ###
If you want to install the development version from [GitHub](https://github.com/oschwery/BoskR), you can do so using the package devtools:
```{r, eval=FALSE}
# 1. Install package 'devtools' (if not already installed):
install.packages("devtools")

# 2. Load 'devtools' and permanently or temporarily install and load 'BoskR':
library(devtools)

dev_mode(on=T)  # if you run this line, you launch developer mode, in which case the package will only be installed temporarily. If you want to install the GitHub version permanently, you can of course just ommit this line.

install_github("oschwery/BoskR", build_vignettes = TRUE)  # install the package from GitHub
library(BoskR)  # load the package

dev_mode(on=F)  # If you used it, leave developers mode after done trying out 'BoskR'
```
Setting `build_vignettes` to `TRUE` ensures that the Rmd file will be knit to an HTML vignette (like the one you're currently reading, probably) that can then be called from R using `vignette("boskr")`. You can ommit this (the default is `FALSE`) if it causes any issues.


## 3 Use ##
When we load the package, all its functions and the other required packages are imported.
```{r, results="hide", warning=FALSE, message=FALSE}
library(BoskR)
```

```{r, echo=FALSE}
set.seed(1337)
library(phyloTop)
```

### 3.1 Load and Prepare Data ###
#### 3.1.1 Example Files ####
The package comes with a set of six empirical example trees to use, which can be loaded using `data()` (for the sake of this demonstration, we want to only use the whales tree, so we extract that one).
```{r, results="hide"}
data(emptesttrees)  # load example treeset
emptesttrees <- emptesttrees$Emp_1  # replace it by first tree only
```

#### 3.1.2 Data Format and Own Data ####
The downstream functions expect the input trees to be in a certain format (object of class `multiPhylo`/list of trees). We can use `CombineTrees()` to combine any of our own input trees, and the function takes anything from single trees, lists of trees and `multiPhylo` objects. Here we can use it to make sure our single whale tree is in the right format as well.
```{r, results="hide"}
emptesttrees <- CombineTrees(emptesttrees)
```

#### 3.1.3 Test and Correct Tree Issues ####
The analyses require rooted, fully resolved, ultrametric trees, and we can test our tree set for that and automatically correct it. If the a tree is not ultrametric due to rounding errors, and if we are fine to randomly resolve polytomies, we can let `TreeCorr()` do that for us (plus reordering all trees to `cladewise`). It won't automatically root unrooted trees, but just tell you which ones are unrooted, so you can root or exclude them.
```{r}
emptesttrees <- TreeCorr(emptesttrees)
```

### 3.2 Estimate Parameters and Get Shape Metrics ###
With our tree ready, we can move on to the main part of the procedure. We first estimate the diversification parameters from our empirical tree under our model of choice (we'll pick `"BD"` for a constant rate birth-death model for now):
```{r, results="hide"}
emptestparams <- GetTreeParams(emptesttrees, "BD")  # estimate parameters from empirical tree
```
Now we can also already get this tree's shape metrics. The calculation of the Laplacian spectra can be a bit time-intensive, so if in a hurry, setting `quick_run` to `TRUE` will skip these and only calculate the other metrics (see below). They can also use quite a bit of memory, thus `return_spectra` is by default set to `FALSE`, which means the actual Laplacian spectra aren't saved, but only the metrics derived from it. Unless you want to dig deeper for some reason, you probably don't need to bother with them anyways.
```
emptestmetrics <- GetTreeMetrics(emptesttrees, empirical_start=TRUE, return_spectra=FALSE, quick_run=FALSE) # get tree shape metrics from empirical trees
```

### 3.3 Simulate Trees Under Model ###
Using the parameters we estimated in the previous step, we now simulate some trees (let's say `100`) under the same model, so we can see what kinds of trees that model produces:
```{r, results="hide", warning=FALSE, message=FALSE}
emptestsims <- GetMetricTreeSets(empirical_start=TRUE, empParams=emptestparams, current_method="BD", Numbsim1=100)
```
Now we get the tree metrics for our freshly simulated trees (we could also use `GetTreeParams()` again to also estimate their parameters, but we'll save that for another day):
```{r, results="hide", warning=FALSE, message=FALSE}
simtestmetrics <- GetTreeMetrics(emptestsims, empirical_start=TRUE, return_spectra=FALSE, quick_run=FALSE)
```

### 3.4 Compare Empirical and Simulated Tree Shapes ###
To see how our simulations compare to the input tree, we can use the distribution of tree metrics to calculate a p-value for the empirical tree's metrics:
```{r}
pvalues <- PvalMetrics(emptestmetrics, simtestmetrics)
pvalues$pValues  # Display table of p-Values
```

For illustration and to get a better idea of what our results look like, we can plot them in a number of different ways. With `plotPvalMetricsPDF` we can see where the empirical metrics fall on the distribution of the simulated ones:
```{r, fig.width=7, fig.height=2}
plotPvalMetricsPDF(emptestmetrics, simtestmetrics, set=1)
```

And with `ScatterMetrics`, we can visualise empirical and simulated metrics in a 3D scatterplot:
```{r, fig.width=5, fig.height=5}
ScatterMetrics(emptestmetrics, simtestmetrics, pair=1, skim=FALSE, combine=FALSE, colours=c("black", "red"))
```

We could also have used `plotPvalMetricsCDF()` to look at the cumulative distribution functions of each metric instead of the probability density function. If we had looked at several trees, we could have used `skim=TRUE` to click through scatterplots of each tree separately, or `combine=TRUE` to get a plot with all empirical and simulated metrics combined.

### 3.5 Adequacy of Models That Aren't Implemented in BoskR ###
Currently, only three models are implemented (constant rate birth-death, time-dependent birth-death, and diversity dependent). However, this approach theoretically works with any model for which we can both estimate rates and simulate trees.

So, if we want to test the adequacy of _e.g._ an environmental-dependent birth-death model, we can simply use that model to estimate the rates of our empirical tree, and use the estimated parameters to simulate an appropriate number of trees under the same model and parameters again.

Using that initial tree and the simulated set, we can use the same procedure as above with the following modifications:

1. Skip `GetTreeParams` and `GetMetricTreeSets`, as we already estimated the parameters under this model and simulated the corresponding tree set
2. Run `CombineTrees()` for the simulated tree set as well, to ensure that it is in the correct input-format for the subsequent functions. To do this for a simulated tree set, set the additional parameter `sims` to `TRUE`. At this point, you have to supply your trees as an object of class `multiPhylo` (essentialy a list of trees, with the class-label set to `multiPhylo`).
   ```{r, results="hide"}
   emptestsims <- CombineTrees(emptestsims, sims=TRUE)
   ```
3. Now go straight to estimating the metrics of both initial tree and simulations separately, and from there follow the standard procedure to get p-values, plot results, etc.

### 3.6 Alternative Metrics ###
Beyond the three main tree shape metrics, _i.e._ Principal Eigenvalue, Asymmetry, and Peakedness of the spectral sensity of the tree's modified graph Laplaican, BoskR also calculates the normalised versions of those. Additionally, it calculates a set of more classical shape metrics (Colless index, Sackin index, number of cherries, number of pitchforks, average ladder size, and the gamma statistic), as well as a set consisting of the minimum, median, and maximum node ages and branch lengths respectively. For an extended exploration of model adequacy and its reasons, those additional sets can also be explored.

All sets are calculated anyway by `GetTreeMetrics`, but the subsequent functions `PvalMetrics` and `plotPvalMetricsPDF` will by default only make use of the standard set (`"spectR"`). The others can be chosen instead when specifying the variable `metricset` accordingly, _i.e._ either `"spectRnorm"`, `"classic"`, or `"nodibranch"` respectively. In case of the p-values, we can also just calculate them `"ALL"`, or all but the spectra ones, if the quick version of the analysis was run (`"AllQuick"`). 
```{r}
pvalues_classic <- PvalMetrics(emptestmetrics, simtestmetrics, metricset="classic")
pvalues_classic$pValues  # Display table of p-Values
```
For reasons of space, we can plot each `metricset` separately, but not all at once.
```{r, fig.width=9, fig.height=2}
plotPvalMetricsPDF(emptestmetrics, simtestmetrics, set=1, metricset="classic")
```

## 4 Final Notes ##
### 4.1 Interpretation and Caveats ###
Following the general idea, if our empirical tree is significantly different from the simulations in one or more of the metrics, or is outside the could of simulated trees in metric space, we should question the adequacy of the employed model.

In order to make sure the results aren't spurious, and in order to generally make sense of them, I would strongly encourage to actually plot a couple of the simulated trees, in order to compare them to the input tree and see how any differences in tree metrics manifest.

### 4.2 Further use and Outlook ###
If one wants to study the behaviour of a model in general, `GetMetricTreeSets()` can also just take parameter values as input, as opposed to an empirical phylogeny.

Another potential use for this approach might be as a way to make sure simulated trees (used _e.g._ to create a distribution of scenarios to compare an empirical solution to) are within a certain range of shape around the empirical example, if the question makes that necessary.

I may want to add more models to the package in the near future, and welcome suggestions as to which to add.
