---
title: "BoskR Tutorial"
author: "[Orlando Schwery](https://oschwery.github.io/)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BoskR Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1 Introduction ##
This is a tutorial vignette to provide a quick and easy introduction to the use of BoskR. It will make use of the inbuilt example files and lead through the most important functions of the package.

The purpose of the package BoskR is, to test whether a given diversification model is adequate to analyse an empirical phylogeny with. We can use different models to analyse diversification on a phylogeny, and pick the one that's best supported ( _e.g._ by likelihood or AIC), but that doesn't tell us anything about whether that model is an appropriate one to use in the first place, _i.e._ whether the model is capable of describing the patterns we're analysing at all.

The way BoskR is trying to judge adequacy is by tree shape. The logic behind this is, that if a model cannot produce trees of a shape similar to an empirical tree, it is also not suited to analyse that tree. The package makes use of tree metrics measured from the spectral density of a trees modified graph Laplacian, as calculated in the package RPANDA. Details on this can be found in the corresponding publications and manuals [add links to those here].

## 2 Installation ##
### 2.1 CRAN ###
In order to install the stable CRAN version of the package:

```{r, eval=FALSE}
install.packages("BoskR")
```
### 2.2 Development Version from GitHub ###
If - for any reason - you would want to install the development version from GitHub, I suggest doing so by installing it temporarily with the help of the package `devtools`:
```{r, eval=FALSE}
# 1. Install package 'devtools' (if not already installed):
install.packages("devtools")

# 2. Load 'devtools' and temporarily install and load 'BoskR' in developers mode:
library(devtools)
dev_mode(on=T)
install_github("oschwery/BoskR")  # install the package from GitHub
library(BoskR)  # load the package

#3. Leave developers mode after done trying out 'BoskR':
dev_mode(on=F)
# the package will not stay on your system permanently like that, which make sense in case of the
# development version, as opposed to the stable version
```

## 3 Use ##

When we load the package, all its functions and the other required packages are imported.
``` {r, results="hide", warning=FALSE, message=FALSE}
library(BoskR)
```

### 3.1 Load Data ###
#### 3.1.1 Example Files ####



##4 Final Notes##









## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output:
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side.

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
