---
title: "BoskR Tutorial"
author: "[Orlando Schwery](https://oschwery.github.io/)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BoskR Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1 Introduction ##
This is a tutorial vignette to provide a quick and easy introduction to the use of BoskR. It will make use of the inbuilt example files and lead through the most important functions of the package.

The purpose of the package BoskR is, to test whether a given diversification model is adequate to analyse an empirical phylogeny with. We can use different models to analyse diversification on a phylogeny, and pick the one that's best supported ( _e.g._ by likelihood or AIC), but that doesn't tell us anything about whether that model is an appropriate one to use in the first place, _i.e._ whether the model is capable of describing the patterns we're analysing at all.

The way BoskR is trying to judge adequacy is by tree shape. The logic behind this is, that if a model cannot produce trees of a shape similar to an empirical tree, it is also not suited to analyse that tree. The package makes use of tree metrics measured from the spectral density of a trees modified graph Laplacian, as calculated in the package `RPANDA`. Details on this can be found in the corresponding publications and manuals [add links to those here].

## 2 Installation ##
### 2.1 CRAN ###
In order to install the stable CRAN version of the package [this is obviously not possible yet]:

```{r, eval=FALSE}
install.packages("BoskR")
```
### 2.2 Development Version from GitHub ###
If you want to install the development version from GitHub, you can do so using the package `devtools`:
```{r, eval=FALSE}
# 1. Install package 'devtools' (if not already installed):
install.packages("devtools")

# 2. Load 'devtools' and permanently or temporarily install and load 'BoskR':
library(devtools)

dev_mode(on=T)  # if you run this line, you launch developer mode, in which case the package will only be installed temporarily.

install_github("oschwery/BoskR")  # install the package from GitHub
library(BoskR)  # load the package

dev_mode(on=F)  # If you used it, leave developers mode after done trying out 'BoskR'
```

## 3 Use ##
When we load the package, all its functions and the other required packages are imported.
``` {r, results="hide", warning=FALSE, message=FALSE}
library(BoskR)
```

### 3.1 Load and Prepare Data ###
#### 3.1.1 Example Files ####

data(emptesttrees)

CombineTrees(whales)


# test and correct
emptesttrees <- TreeCorr(emptesttrees)

### 3.2 Estimate Parameters and Get Shape Metrics ###

# estimate parameters from empirical trees
emptestparams <- GetTreeParams(emptesttrees, curmeth)
# get tree shape metrics from empirical trees
emptestmetrics <- GetTreeMetrics(emptesttrees, empirical_start=TRUE)

### 3.3 Simulate Trees Under Model ###

# simulate trees based on empirical parameters
emptestsims <- GetMetricTreeSets(empirical_start=TRUE, empParams=emptestparams, current_method=curmeth, Numbsim1=sims)

### 3.4 Get Metrics (and Parameters) of Simulated Trees ###

# estimate parameters and tree shape metrics from simulated trees
simtestmetrics <- GetTreeMetrics(emptestsims, empirical_start=TRUE)
simtestparams <- GetTreeParams(emptestsims, curmeth)

### 3.5 Get Metrics (and Parameters) of Simulated Trees ###

# get p-values for each metrics for all trees
pvalues <- PvalMetrics(emptestmetrics, simtestmetrics)

### 3.6 Plot Results ###

# plot pvalues on CDF and PDF
plotPvalMetricsCDF(pvalues, set=NULL)
plotPvalMetricsPDF(emptestmetrics, simtestmetrics, set=NULL)

# plot 3D scatterplot of single pair of emp and sim tree metrics
ScatterMetrics(emptestmetrics, simtestmetrics, pair=1, skim=FALSE, colours=c("black", "red"))
# plot 3D scatterplot of all pairs of emp and sim metrics, forwarding on prompt
ScatterMetrics(emptestmetrics, simtestmetrics, pair=1, skim=TRUE, colours=c("black", "red"))
# plot 3D scatterplot of ALL emp and sim Metrics combined
ScatterMetrics(emptestmetrics, simtestmetrics, pair=1, combine=TRUE, colours=c("black", "red"))



##4 Final Notes##









## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output:
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side.

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
