---
title: "BoskR Tutorial"
author: "[Orlando Schwery](https://oschwery.github.io/)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BoskR Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1 Introduction ##
This is a tutorial vignette to provide a quick and easy introduction to the use of BoskR. It will make use of the inbuilt example files and lead through the most important functions of the package.

The purpose of BoskR is, to test whether a given diversification model is adequate to analyse an empirical phylogeny with. We can use different models to analyse diversification on a phylogeny, and pick the one that's best supported ( _e.g._ by likelihood or AIC), but that doesn't tell us anything about whether that model is an appropriate one to use in the first place, _i.e._ whether the model is capable of describing the patterns we are attempting to analyse at all.

The way BoskR is trying to judge adequacy is by tree shape. The logic behind this is, that if a model cannot produce trees of a shape similar to an empirical tree, it is also not suited to analyse that tree. The package makes use of tree metrics measured from the spectral density of a tree's modified graph Laplacian, as obtained through the package RPANDA. Details on this can be found in the corresponding publication^[Eric Lewitus, Helene Morlon; Characterizing and Comparing Phylogenies from their Laplacian Spectrum, Systematic Biology, Volume 65, Issue 3, 1 May 2016, Pages 495–507, [https://doi.org/10.1093/sysbio/syv116](https://doi.org/10.1093/sysbio/syv116)] and package documentations^[[Package RPANDA on CRAN](https://CRAN.R-project.org/package=RPANDA)].

## 2 Installation ##
### 2.1 CRAN ###
To just use the stable out of the box version of BoskR, install the stable CRAN version of the package:

```{r, eval=FALSE}
install.packages("BoskR")
```

### 2.2 Development Version from GitHub ###
If you want to install the development version from [GitHub](https://github.com/oschwery/BoskR), you can do so using the package devtools:
```{r, eval=FALSE}
# 1. Install package 'devtools' (if not already installed):
install.packages("devtools")

# 2. Load 'devtools' and permanently or temporarily install and load 'BoskR':
library(devtools)

dev_mode(on=T)  # if you run this line, you launch developer mode, in which case the package will only be installed temporarily.

install_github("oschwery/BoskR")  # install the package from GitHub
library(BoskR)  # load the package

dev_mode(on=F)  # If you used it, leave developers mode after done trying out 'BoskR'
```

## 3 Use ##
When we load the package, all its functions and the other required packages are imported.
```{r, results="hide", warning=FALSE, message=FALSE}
library(BoskR)
```

```{r, echo=FALSE}
set.seed(1337)
```

### 3.1 Load and Prepare Data ###
#### 3.1.1 Example Files ####
The package comes with a set of six empirical example trees to use, which can be loaded using `data()` (for the sake of this demonstration, we want to only use the whales tree, so we extract that one).
```{r, results="hide"}
data(emptesttrees)  # load example treeset
emptesttrees <- emptesttrees$Emp_1  # replace it by first tree only
```

#### 3.1.2 Data Format and Own Data ####
The downstream functions expect the input trees to be in a certain format (object of class `multiPhylo`/list of trees). We can use `CombineTrees()` to combine any of our own input trees, and the function takes anything from single trees, lists of trees and `multiPhylo` objects. Here we can use it to make sure our single whale tree is in the right format as well.
```{r, results="hide"}
emptesttrees <- CombineTrees(emptesttrees)
```

#### 3.1.3 Test and Correct Tree Issues ####
The analyses require fully resolved, ultrametric trees, and we can test our tree set for that and automatically correct it. If the a tree is not ultrametric due to rounding errors, and if we are fine to randomly resolve polytomies, we can let `TreeCorr()` do that (plus reordering all trees to `cladewise`) for us.
```{r}
emptesttrees <- TreeCorr(emptesttrees)
```

### 3.2 Estimate Parameters and Get Shape Metrics ###
With our tree ready, we can move on to the main part of the procedure. We first estimate the diversification parameters from our empirical tree under our model of choice (we'll pick `"BD"` for a constant rate birth-death model for now), and measure the tree's shape metrics:
```{r, results="hide"}
emptestparams <- GetTreeParams(emptesttrees, "BD")  # estimate parameters from empirical tree
emptestmetrics <- GetTreeMetrics(emptesttrees, empirical_start=TRUE) # get tree shape metrics from empirical trees
```

### 3.3 Simulate Trees Under Model ###
Using the parameters we estimated in the previous step, we now simulate some trees (let's say `100`) under the same model, so we can see what kinds of trees that model produces:
```{r, results="hide", warning=FALSE, message=FALSE}
emptestsims <- GetMetricTreeSets(empirical_start=TRUE, empParams=emptestparams, current_method="BD", Numbsim1=100)
```
Now we get the tree metrics for our freshly simulated trees (we could also use `GetTreeParams()` again to also estimate their parameters, but we'll save that for another day):
```{r, results="hide", warning=FALSE, message=FALSE}
simtestmetrics <- GetTreeMetrics(emptestsims, empirical_start=TRUE)
```

### 3.4 Compare Empirical and Simulated Tree Shapes ###
To see how our simulations compare to the input tree, we can use the distribution of tree metrics to calculate a p-value for the empirical tree's metrics:
```{r}
pvalues <- PvalMetrics(emptestmetrics, simtestmetrics)
pvalues$pValues  # Display table of p-Values
```
For illustration and to get a better idea of what our results look like, we can plot them in a number of different ways. With `plotPvalMetricsPDF` we can see where the empirical metrics fall on the distribution of the simulated ones:
```{r, fig.width=7, fig.height=2}
plotPvalMetricsPDF(emptestmetrics, simtestmetrics, set=1)
```

And with `ScatterMetrics`, we can visualise empirical and simulated metrics in a 3D scatterplot:
```{r, fig.width=5, fig.height=5}
ScatterMetrics(emptestmetrics, simtestmetrics, pair=1, skim=FALSE, combine=FALSE, colours=c("black", "red"))
```

We could also have used `plotPvalMetricsCDF()` to look at the cumulative distribution functions of each metric instead of the probability density function. If we had looked at several trees, we could have used `skim=TRUE` to click through scatterplots of each tree separately, or `combine=TRUE` to get a plot with all empirical and simulated metrics combined.

## 4 Final Notes ##
### 4.1 Interpretation and Caveats ###
Following the general idea, if our empirical tree is significantly different from the simulations in one or more of the metrics, or is outside the could of simulated trees in metric space, we should question the adequacy of the employed model.

In order to make sure the results aren't spurious, and in order to generally make sense of them, I would strongly encourage to actually plot a couple of the simulated trees, in order to compare them to the input tree and see how any differences in tree metrics manifest.

### 4.2 Further use and Outlook ###
If one wants to study the behaviour of a model in general, `GetMetricTreeSets()` can also just take parameter values as input, as opposed to an empirical phylogeny.

Another potential use for this approach might be as a way to make sure simulated trees (used _e.g._ to create a distribution of scenarios to compare an empirical solution to) are within a certain range of shape around the empirical example, if the question makes that necessary.

Currently, only three models are implemented (constant rate birth-death, time-dependent birth-death, and diversity dependent), while this theoretically works with any model for which we can both estimate rates and simulate trees. I intend to add more models to the package in the near future, and welcome suggestions as to which to add.
